Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-16 csf-2017/paper.tex
Pass over LR section done

The \emph{observation relation} $\observations$ defines configurations that we
consider safe. Roughly speaking, a configuration is safe if, whenever it halts,
there is a private future world that the resulting memory satisfies. Notice that
failing is considered safe behavior. In fact, the machine often resorts to
failing when an unauthorized access is attempted, like loading from a capability
without read permission. This is similar to \cite{Devriese:2016ObjCap}'s logical
relation for an untyped language, but unlike typical logical relations
for typed languages, which require that programs do not fail.

@@

For a capability with write permission, $\writeCond{}$ must be satisfied for the
capability's range of authority. Such a capability may be used to write any
word, so the region governing the relevant memory must (at least) allow any safe
word to be stored. More precisely, for permit-write-local capabilities it needs
to allow arbitrary safe values to be stored, while for write capabilities,
storing values that are both non-local and safe is sufficient. Like before, we
use $\var{localityReg}$ to pick an appropriate region for the capability's
locality. The condition for the region to be address-stratified is a technical
requirement, defined in the technical appendix. In the value relation, the write
condition is used with $\iota^\pwl$ for write-local capabilities. For write
capabilities, it is used with another standard region
$\iota^\nwl_{\start,\addrend}$ which requires that the words in
$[\start,\addrend]$ are non-local and safe.

@@

\subsection{Safety of the capability machine}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalises the guarantees offered by
the capability machine. Essentially, it says that a capability that only grants
safe authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{\footnotesize{ and }}\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{\footnotesize{ and }} \\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
The Fundamental Theorem can be understood as a general expression of the
guarantees offered by the capability machine, an instance of a general property
called capability safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}. To
understand this, consider that the theorem says that the capability
$((\perm,\gl),\start,\addrend,\addr)$ is safe as a program counter, without any
assumption about what instructions it actually points to (the only assumptions
we have are about the read or write authority that it carries). As such, the
theorem expresses the capability safety of the machine, which guarantees that
\emph{any} instruction is fine and will not be able to go beyond the authority
of the values it has access to.

>>>>

The \emph{observation relation} $\observations$ defines configurations that we
consider safe. Roughly speaking, a configuration is safe if whenever it halts,
there is a private future world that the resulting memory satisfies. Notice that
failing is considered safe behavior. In fact, the machine often resorts to
failing when an unauthorized access is attempted, like loading from a capability
without read permission. This is similar to \cite{Devriese:2016ObjCap}'s logical
relation for an untyped language, but unlike typical logical relations
for typed languages, which require that programs do not fail.

@@

For a capability with write permission, $\writeCond{}$ must be
satisfied for the capability's range of authority. Such a capability
may be used to write any word, so the region governing the relevant
memory must (at least) allow any safe word to be stored. More
precisely, for write-local capabilities it needs to allow arbitrary
safe values to be stored, while for write capabilities, storing values
that are both non-local and safe is sufficient.  This is done in the
value relation by the regions the write condition is applied with. For
write-local capabilities $\iota^\pwl$ is used, and for write
capabilities we use another standard region
$\iota^\nwl_{\start,\addrend}$ which requires that the words in
$[\start,\addrend]$ are non-local and safe.  Like before, we use
$\var{localityReg}$ to pick an appropriate region for the capability's
locality. The condition for the region to be address-stratified is a
technical requirement, defined in the technical appendix.

@@

\subsection{Safety of the capability machine}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalises the guarantees offered by
the capability machine. Essentially, it says that a capability that only grants
safe authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{\footnotesize{ and }}\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{\footnotesize{ and }} \\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
The Fundamental Theorem can be understood as a general expression of the
guarantees offered by the capability machine, an instance of a general property
called capability safety~\citep{Devriese:2016ObjCap,Maffeis2010OC}. To
understand this, consider that the theorem says the capability
$((\perm,\gl),\start,\addrend,\addr)$ is safe as a program counter, without any
assumption about what instructions it actually points to (the only assumptions
we have are about the read or write authority that it carries). As such, the
theorem expresses the capability safety of the machine, which guarantees that
\emph{any} instruction is fine and will not be able to go beyond the authority
of the values it has access to.
